---
title: "Homework 3"
author: Laura Bulmer
date: 10/16/2024
output: github_document
---
```{r}
library(p8105.datasets)
library(tidyverse)
library(ggridges)
library(patchwork)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Problem 1

```{r loading dataset}
data("ny_noaa")
```

## Problem 2

First, the below code is used to load and tidy the datasets related to accelerometers and the participants. 

```{r load and tidy participants}
participants = 
  read_csv(file = "./data/nhanes_covar.csv", 
           skip = 4,
           na = c(".", "NA", ""),
           col_types = cols(
             'SEQN' = col_character(),
             'sex' = col_factor(),
             'age' = col_integer(),
             'BMI' = col_double(),
             'education' = col_factor()
           ))

participants = janitor::clean_names(participants)
```


```{r load and tidy accelerometer}
accelerometer = read_csv(file = "./data/nhanes_accel.csv",
                         na = c(".", "NA", ""),
                         col_types = cols(
                           'SEQN' = col_character()
                         )
                         )
accelerometer = janitor::clean_names(accelerometer)
```

Next I am sub-setting participants data to drop missing observations and exclude participants less than 21.

```{r }
participants_clean = participants %>%
  drop_na() %>%
  filter(age > 20)
```

Next, I am joining my two datasets. 

```{r}
accel_df = left_join(participants_clean, accelerometer, by = 'seqn')
```

Next I am creating a table for the number of men and women in each education category.

```{r}
accel_df %>%
  mutate(
    sex = recode(sex, `1` = "Male", `2` = "Female"),
    education = recode(education, 
                       `1` = "Less than High School", 
                       `2` = "High School Equivalent", 
                       `3` = "More than High School")
    )%>%
group_by(education, sex) %>%
  summarize(count_education = n()) %>%
  pivot_wider(
    names_from = sex,
    values_from = count_education
  )%>%
knitr::kable()
```

```{r}
accel_df %>%
  ggplot(aes(x = education, y = age, fill = sex))+
  geom_boxplot()+
  labs(title = "Age Distribution of Men and Women by Education Category",
       x = "Education Category",
       y = "Age")
```


INSERT COMMENTS ON THE TABLE AND PLOT ABOVE HERE

Next I am aggregating and plotting total activity.

```{r}
accel_df = accel_df %>%
  mutate(total_act = rowSums(select(., min1:min1440), na.rm = TRUE))

accel_df%>%
  ggplot(aes(x = age, y = total_act, color = sex)) +
    geom_point() +  
    geom_smooth(method = "lm", se = FALSE) +
    facet_wrap(~ education) + 
    labs(title = "Total Activity by Age and Education Level",
         x = "Age",
         y = "Total Activity",
         color = "Sex")
```

